---
title: "Quad versus Tibialis"
author: "Adam Struck"
date: "February 25, 2015"
output:
  pdf_document:
    highlight: tango
    toc: yes
    toc_depth: 3
---

```{r Configuration, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}
library(knitr)
library(ggplot2)
library(topGO)
library(org.Hs.eg.db)
library(dplyr)
library(tidyr)
library(data.table)
library(foreach)
library(stringr)
library(Rgraphviz)

opts_chunk$set(background='gray80', echo = TRUE, tidy=TRUE, 
               warning = FALSE, cache=FALSE, comment='', dpi=72)
```

```{r Functions}
find.dm.events <- function(DT) {
    if("gene_symbol" %in% colnames(DT)){
        DT %>%
        filter(abs(delta_psi_mean) >= 0.05,
               Control_n / max(Control_n, na.rm = TRUE)  >= 0.75,
               DM1_n / max(DM1_n, na.rm = TRUE) >= 0.75,
               DM1_n_sig / DM1_n >= 0.25) %>%
        select(gene_symbol, event_name, isoforms, Control_psi_mean, Control_psi_sd, Control_n,
                      DM1_psi_mean, DM1_psi_sd, DM1_n, delta_psi_mean, DM1_n_sig) %>%
        arrange(desc(abs(delta_psi_mean)))
    } else {
        DT %>%
        filter(abs(delta_psi_mean) >= 0.05,
               Control_n / max(Control_n, na.rm = TRUE)  >= 0.75,
               DM1_n / max(DM1_n, na.rm = TRUE) >= 0.75,
               DM1_n_sig / DM1_n >= 0.25) %>%
        select(event_name, isoforms, Control_psi_mean, Control_psi_sd, Control_n,
                      DM1_psi_mean, DM1_psi_sd, DM1_n, delta_psi_mean, DM1_n_sig) %>%
        arrange(desc(abs(delta_psi_mean)))
    }
}
```

## Identify mis-regulated nonUTRevents
```{r Load data}
event_type <- "nonUTRevents.multi"

## Load healthy quadricep versus tibialis results
allControls_res <- tbl_dt(fread(paste("~/Projects/DMseq/results/allControls/allControls", event_type, "results.txt", sep = "_")))
allControls_res <- allControls_res %>% mutate(delta_psi = Quad_psi_mean - Tibialis_psi_mean)

## filter to identify events with differnet splicing patterns between the tissues
quad_vs_tibialis <- allControls_res %>%
    select(gene_symbol, event_name, Quad_psi_mean, Quad_n, Tibialis_psi_mean, 
    			 Tibialis_n, Quad_vs_Tibialis_n_sig, isoforms, delta_psi) %>%
            filter(Quad_n / max(Quad_n, na.rm = TRUE) >= 0.75,
                   Tibialis_n / max(Tibialis_n, na.rm = TRUE) >= 0.75,
                   abs(delta_psi) >= 0.05, Quad_vs_Tibialis_n_sig / Quad_n >= 0.25) %>%
                       arrange(desc(Quad_vs_Tibialis_n_sig))

## Load tibialis & quadricep DM_vs_Control results
tibialis_pdata <- tbl_dt(fread("~/Projects/DMseq/data/DM_tibialis_pdata.txt"))
tibialis_res <- tbl_dt(fread(paste("~/Projects/DMseq/results/tibialis/tibialis", event_type, "results.txt", sep = "_")))
tibialis_res <- tibialis_res %>% mutate(frac_sig = DM1_n_sig / DM1_n)

quadricep_pdata <- tbl_dt(fread("~/Projects/DMseq/data/DM_quadricep_pdata.txt"))
quadricep_res <- tbl_dt(fread(paste("~/Projects/DMseq/results/quadricep/quadricep", event_type, "results.txt", sep = "_")))
quadricep_res <- quadricep_res %>% mutate(frac_sig = DM1_n_sig / DM1_n)

## Filter to identify events mis-regulated in DM1
dm_tibialis <- find.dm.events(tibialis_res)
dm_quadricep <- find.dm.events(quadricep_res)
```

## Compare delta psi values
```{r}
event_set <- intersect(union(dm_tibialis$isoforms, dm_quadricep$isoforms), quad_vs_tibialis$isoforms)

deltapsi_data <- Reduce(function(...) merge(..., by = "isoforms", all = TRUE),
                        list(select(allControls_res, gene_symbol, event_name, isoforms, delta_psi) %>% filter(isoforms %in% event_set),
                             select(tibialis_res, isoforms, delta_psi_mean) %>% filter(isoforms %in% event_set),
                             select(quadricep_res, isoforms, delta_psi_mean) %>% filter(isoforms %in% event_set)))
setnames(deltapsi_data, c("isoforms", "gene_symbol", "event_name", "Quad_vs_Tibialis_deltapsi", "Tibialis", "Quad"))
```

## Scatterplots of delta psi values
### All events dyesregulated in either quad or tibialis (DM vs Control)

```{r}
event_set <- union(dm_tibialis$isoforms, dm_quadricep$isoforms)
deltapsi_data <- Reduce(function(...) merge(..., by = "isoforms", all = TRUE),
                        list(select(allControls_res, gene_symbol, event_name, isoforms, delta_psi) %>% filter(isoforms %in% event_set),
                             select(tibialis_res, isoforms, delta_psi_mean) %>% filter(isoforms %in% event_set),
                             select(quadricep_res, isoforms, delta_psi_mean) %>% filter(isoforms %in% event_set)))
setnames(deltapsi_data, c("isoforms", "gene_symbol", "event_name", "Quad_vs_Tibialis_deltapsi", "Tibialis", "Quad"))

ggplot(deltapsi_data, aes(x = Tibialis, y = Quad, colour = cut(deltapsi_data$Quad_vs_Tibialis_deltapsi, seq(-1,1, 0.1)))) +
    geom_point() +
        geom_abline(yintercept = 0, slope = 1, linetype = "longdash") +
            labs(y = "Quad Delta PSI \n DM versus Control",
                 x = "Tibialis Delta PSI \n DM versus Control") +
                     scale_color_discrete(name = "Quad_vs_Tibialis_deltapsi")
```

### Events dyesregulated in either quad or tibialis (DM vs Control) that are also differentially spliced between healthy quad and healthy tibialis

```{r}
event_set <- intersect(union(dm_tibialis$isoforms, dm_quadricep$isoforms), quad_vs_tibialis$isoforms)
deltapsi_data <- Reduce(function(...) merge(..., by = "isoforms", all = TRUE),
                        list(select(allControls_res, gene_symbol, event_name, isoforms, delta_psi) %>% filter(isoforms %in% event_set),
                             select(tibialis_res, isoforms, delta_psi_mean) %>% filter(isoforms %in% event_set),
                             select(quadricep_res, isoforms, delta_psi_mean) %>% filter(isoforms %in% event_set)))
setnames(deltapsi_data, c("isoforms", "gene_symbol", "event_name", "Quad_vs_Tibialis_deltapsi", "Tibialis", "Quad"))

ggplot(deltapsi_data, aes(x = Tibialis, y = Quad, colour = cut(deltapsi_data$Quad_vs_Tibialis_deltapsi, seq(-1,1, 0.1)))) +
    geom_point() +
        geom_abline(yintercept = 0, slope = 1, linetype = "longdash") +
            labs(y = "Quad Delta PSI \n DM versus Control",
                 x = "Tibialis Delta PSI \n DM versus Control") +
                     scale_color_discrete(name = "Quad_vs_Tibialis_deltapsi")

deltapsi_data_long <- gather(deltapsi_data, key = "Tissue", value = "DM_vs_Control_deltapsi", 5:6)
ggplot(deltapsi_data_long, aes(x = Quad_vs_Tibialis_deltapsi, y = DM_vs_Control_deltapsi, colour = Tissue)) +
    geom_point() +
        geom_hline(yintercept = 0, linetype = "longdash") +
            geom_vline(xintercept = 0, linetype = "longdash") +
                labs(y = "Delta PSI \n DM versus Control",
                     x = "Delta PSI \n Control Quad versus Control Tibialis")
```

### Linear model residuals [DM_Tibialis_deltapsi ~ DM_Quad_deltapsi] plotted against healthy quad_vs_tibialis deltapsi

```{r}
fit <- lm(Tibialis ~ Quad, data = deltapsi_data)
cor.test(fit$residuals, deltapsi_data$Quad_vs_Tibialis_deltapsi)

ggplot() +
    geom_point(aes(y = deltapsi_data$Quad_vs_Tibialis_deltapsi, x = fit$residuals)) +
            labs(x = "Residuals \n lm(DM_Tibialis_deltapsi ~ DM_Quad_deltapsi)",
                 y = "Delta PSI \n Quad versus Tibials")
```

## Gene ontology term enrichment for dysregulated events in quad OR tibialis (DM vs Control) AND between controls (Quad vs Tib)
```{r, results='hide', warning=FALSE}
all_muscle_genes <- filter(allControls_res, Quad_n / max(Quad_n, na.rm = TRUE) >= 0.75,                                                                                                                                                        
                           Tibialis_n / max(Tibialis_n, na.rm = TRUE) >= 0.75)$gene_symbol %>%                                                                                                                                                 
                               unique() 

## different in quad OR tibialis (DM vs Control) AND between controls (Quad vs Tib)
sig_genes <- intersect(union(dm_tibialis$gene_symbol, dm_quadricep$gene_symbol), quad_vs_tibialis$gene_symbol)

ontology_class <- "BP"

myGO2genes <- AnnotationDbi::select(org.Hs.eg.db, keys = all_muscle_genes, columns=c("ENSEMBL", "GO"), keytype="SYMBOL")
myGO2genes <- myGO2genes %>% filter(!is.na(ENSEMBL)) %>% tbl_df
myGO2genesList <- tapply(filter(myGO2genes, ONTOLOGY == ontology_class)$ENSEMBL, filter(myGO2genes, ONTOLOGY == ontology_class)$GO, FUN = c)

ensemblIDs <- myGO2genes$ENSEMBL[match(all_muscle_genes, myGO2genes$SYMBOL)]
geneList <- factor(as.integer(all_muscle_genes %in% sig_genes))
names(geneList) <- ensemblIDs

GOdata <- new("topGOdata", description = "GO analysis of genes with differential splicing",
              ontology = ontology_class, allGenes = geneList, nodeSize = 5,
              annot = annFUN.GO2genes, GO2genes = myGO2genesList)

resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
resultFisher.elim <- runTest(GOdata, algorithm = "elim", statistic = "fisher")

allRes <- GenTable(GOdata, classicFisher = resultFisher,
                   elimFisher = resultFisher.elim,
                   orderBy = "elimFisher", ranksOf = "elimFisher",
                   topNodes = 20)
```

```{r}
head(sig_genes, n = 50)
```

## Ontology term enrichment results
```{r}
kable(allRes)

showSigOfNodes(GOdata, score(resultFisher.elim), firstSigNodes = 10, useInfo = "all")
```
